---
title: "Gender Bias in Promotions: Associate to Full Professor"
author: "Andrew Jumanca"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(boot)
```

## Data Preparation
This dataset tracks faculty members ranks over time to assess whether gender bias exists in promotions from Associate to Full Professor.

```{r}
data <- read.csv("salary_csv.txt")

# Filter to Associate Professors only, tracking promotion to Full Professor
data <- data %>% 
  filter(rank %in% c("Assoc", "Full")) %>%
  group_by(id) %>%
  mutate(promoted = as.integer("Full" %in% rank & "Assoc" %in% rank)) %>%
  ungroup()
```

## Logistic Regression: Predicting Promotion Probability
Logistic regression is appropriate because promotion is a binary outcome. This model assesses the impact of gender while controlling for other factors.

```{r}
logit_model <- glm(promoted ~ sex + field + admin + deg + salary, data = data, family = binomial)
summary(logit_model)
```

## Two-Sample Z-Test: Promotion Rates by Gender
A Z-test is useful for comparing proportions between two independent groups. Here, we compare the promotion rates of male and female faculty members.

```{r}
n_f <- sum(data$sex == "F")
n_m <- sum(data$sex == "M")
p_f <- mean(data$promoted[data$sex == "F"])
p_m <- mean(data$promoted[data$sex == "M"])
se <- sqrt((p_f * (1 - p_f) / n_f) + (p_m * (1 - p_m) / n_m))
z_score <- (p_f - p_m) / se
p_value <- 2 * pnorm(-abs(z_score))
list(z_score = z_score, p_value = p_value)
```

## Bootstrapping: Gender and Promotion Rate
Bootstrapping resamples the dataset to estimate the variability in promotion rates. This method is useful when making inferences about small sample sizes.

```{r}
promotion_rate <- function(data, indices) {
  d <- data[indices, ]
  mean(d$promoted[d$sex == "F"]) - mean(d$promoted[d$sex == "M"])
}
boot_res <- boot(data, promotion_rate, R = 1000)
boot_res$t0
```

## ANOVA: Gender & Other Factors on Salary
ANOVA tests for differences in means across multiple groups. Here, we assess whether gender and other factors significantly affect salary.

```{r}
aov_model <- aov(salary ~ sex + field + admin + deg, data = data)
summary(aov_model)
```

## Trends Over Time: Promotion Rates by Decade
Visualizing trends over time helps determine whether gender disparities in promotion have changed across different decades.

```{r}
data <- data %>% mutate(decade = floor((year + 1900) / 10) * 10)
ggplot(data, aes(x = as.factor(decade), fill = sex)) +
  geom_bar(position = "fill") +
  labs(title = "Promotion Rates by Decade and Gender", y = "Proportion Promoted", x = "Decade")
```
