---
title: "Project Milestone 2"
author: "John Tilelli"
date: "2025-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


setwd("/Users/johntilelli/Library/Mobile Documents/com~apple~CloudDocs/UW MSDS/DATA 557/Course project ")
```

```{r load_data}

#read raw data
data <- readLines("salary.txt")

data_trim <- trimws(data)  # Remove leading/trailing spaces & tabs


#clean file and ensure consistent separators
data_csv <- gsub("\\s+", ",", data_trim)  # Convert spaces & tabs to commas

#write new file
writeLines(data_csv, "salary_csv.txt")

data_cleaned <- read.csv("salary_csv.txt", header=TRUE, stringsAsFactors=FALSE)
head(data_cleaned)

#validation
#check for duplicates in ID column
duplicates <- data_cleaned %>%
  group_by(id) %>%
  filter(n() > 1) %>%
  select(id) %>%
  distinct()

```

**Prompt:** 1. Generate suitable descriptive statistics of the distribution of academic field, year in which they attained their highest degree, year they were hired at the university, academic rank in 1995, and monthly salary in 1995 by sex.

**Notes:**
academic field - categorical, counts per category
year in which they attained their highest degree 
year they were hired at the university
academic rank in 1995
monthly salary in 1995 by sex

to-do
cross-tabs for m/f

**Response:** 

**Code:**
```{r}
summary(data_cleaned)
```

**Academic field:**


```{r acamdemic_field}



```

```{r}
summary(data_cleaned)

#academic field
library(dplyr)
field_counts <- data_cleaned %>%
  count(field, name = "count") %>%
  arrange(desc(count))  #sort

print(field_counts)

library(ggplot2)
ggplot(field_counts, aes(x = reorder(field, -count), y = count)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Cases by Academic Field", x = "Academic Field", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate labels for readability


#Year of Highest Degree Attained
hist(data_cleaned$yrdeg, 
     main="Histogram of Year of Highest Degree Attained", 
     xlab="Year Degree Attained", 
     col="skyblue", 
     border="black")

library(ggplot2)

# Create a histogram with different colors for Male and Female
ggplot(data_cleaned, aes(x=yrdeg, fill=sex)) +
  geom_histogram(binwidth=5, position="identity", alpha=0.6, color="black") +
  labs(title="Histogram of Year of Highest Degree Attained", 
       x="Year Degree Attained", 
       y="Frequency") +
  scale_fill_manual(values=c("blue", "red")) +  # Male = Blue, Female = Red
  theme_minimal()



#year they were hired at the university
hist(data_cleaned$startyr, 
     main="Histogram of Year of Highest Degree Attained", 
     xlab="Year hired at university", 
     col="skyblue", 
     border="black")

rank_counts <- data_cleaned %>%
  count(rank, name = "count") %>%
  arrange(desc(count))  #sort

library(ggplot2)
ggplot(rank_counts, aes(x = reorder(rank, -count), y = count)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Cases by Rank", x = "Academic Field", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

salary_male <- data_cleaned %>% filter(sex == "M") %>% pull(salary)
salary_female <- data_cleaned %>% filter(sex == "F") %>% pull(salary)


#histogram of salaries by male and female
hist(salary_male, 
     main="Salary Distribution (1995) - Male", 
     xlab="Salary", 
     col="blue", 
     border="black")

hist(salary_female, 
     main="Salary Distribution (1995) - Female", 
     xlab="Salary", 
     col="red", 
     border="black")


```

**Prompt:** 2. Write a paragraph commenting on any differences you observe between men and women faculty in these measurements. 

**Notes:**
summarize data loading process
take note on the granularity of the data
validation of load - 
max
min
counts

process - we will each do our own descriptive analysis and then bring the results together for tomorrow and validate our results. 

```{r}
```
