## Group 12 Milestone 2

#### Descriptive Statistics:

```{r echo=FALSE, message=FALSE}
df <- read.csv("../salary_csv.txt", header=TRUE, stringsAsFactors=FALSE)
head(df)
```

```{r echo=FALSE, warning = FALSE, message=FALSE}
library(ggplot2)
library(dplyr)

# Count total number of M and F in the dataset
total_sex <- df %>% count(sex)

for (stat in c("field", "yrdeg", "startyr", "rank", "salary")) {
  if (stat %in% c("yrdeg", "startyr", "salary")) {

    df <- df %>%
      mutate(binned = cut(!!sym(stat), breaks = 5, dig.lab = 5))
    df %>%
      count(binned, sex) %>%
      left_join(total_sex, by = "sex", suffix = c("", "_total")) %>%
      mutate(prop = n / n_total * 100) %>%
      ggplot(aes(x = binned, y = prop, fill = sex)) +
      geom_bar(stat = "identity", position = position_dodge2(preserve = "single")) +
      geom_text(aes(label = sprintf("%.2f%%", prop)), position = position_dodge2(width = 0.9, preserve = "single"), vjust = -0.5) +
      labs(x = stat, y = "Proportion (%)", fill = "Sex") +
      theme_minimal() +
      ggtitle(paste("Proportion of", stat, "by Sex")) -> p
    print(p)
  } else {

    df %>%
      count(!!sym(stat), sex) %>%
      left_join(total_sex, by = "sex", suffix = c("", "_total")) %>%
      mutate(prop = n / n_total * 100) %>%
      ggplot(aes(x = !!sym(stat), y = prop, fill = sex)) +
      geom_bar(stat = "identity", position = position_dodge2(preserve = "single")) +
      geom_text(aes(label = sprintf("%.2f%%", prop)), position = position_dodge2(width = 0.9, preserve = "single"), vjust = -0.5) +
      labs(x = stat, y = "Proportion (%)", fill = "Sex") +
      theme_minimal() +
      ggtitle(paste("Proportion of", stat, "by Sex")) -> p
    print(p)
  }
}
```

#### Salary distributions evaluated by sex, separately
```{r echo=FALSE}
salary_male <- data_cleaned %>% filter(sex == "M" & year == 95) %>% pull(salary)
salary_female <- data_cleaned %>% filter(sex == "F"& year == 95) %>% pull(salary)
 
hist(salary_male, 
     main="Salary Distribution (1995) - Male", 
     xlab="Salary", 
     col="blue", 
     border="black")
 
hist(salary_female, 
     main="Salary Distribution (1995) - Female", 
     xlab="Salary", 
     col="red", 
     border="black")
```

### Analysis
The above analysis is meant to serve as a starting point before our team dives deeper into more meaningful statistical 
analysis to determine whether or not there is a difference in salary between male and female college professors. When 
splitting the data by field, 'other' accounts for nearly 70% of both male and female college professors. 20% of females 
and 12% of males are in the 'arts' field and 10% of females and 22% of males are in the 'prof' field. When looking at the 
year in which each professor attained their highest degree, the distribution for males and females is quite similar and 
normally distributed around 1968 for males and 1977 for females. When it comes to start year, women start quite a bit 
later than men in this data set. The male's average start year is 1975, while the female's is 1980. Looking at rank, 
when grouped by sex, is quite interesting. Slightly more than two-thirds of females are either assistant or associate 
professors, while only 50% of males are in one of these two ranks. The other half of men are full professors, while 
only 25% of females are full professors. Finally, it appears that males tend to make more than females in this data 
set, when not controlling for any factors. The top quartile of females begins at $4951 monthly, compared to that of 
the males which begin at $6040 monthly.
