
Question 2: Has sex bias existed in the starting salaries of faculty members (salaries in the year hired)?

```{r}
df <- read.table("SalaryAnalysis/salary.txt", header = TRUE, sep = "", stringsAsFactors = FALSE)
```

```{r}
df_male <- subset(df, sex == "M")
df_female <- subset(df, sex == "F")
```

Filter by start year
```{r}
df_starting <- df[df$year == df$startyr, ]
```

Create Dist from Degree Variable 
```{r}
df_starting$dist_from_degree <- df_starting$startyr - df_starting$yrdeg
```

Examine Difference Between Men and Women Salaries Using Box Plots 
```{r}
library(ggplot2)

ggplot(df_starting, aes(x = sex, y = salary, fill = sex)) +
  geom_boxplot() +
  labs(title = "Starting Salary Distribution by Sex",
       x = "Sex",
       y = "Starting Salary") +
  scale_fill_manual(values = c("blue", "pink")) +  # Customize colors
  theme_minimal()
```
Far more outliers on the M side of things, but similar means and distributions during starting years. 

Null Hypothesis: There is no difference between the salaries of men and women in starting year
Alternative Hypothesis: There is a difference between the salaries of men and women in starting year 
```{r}
t.test(salary ~ sex, data = df_starting, var.equal = TRUE) 
```
Based on a two sample t-test, reject the null hypothesis that there is no difference in salary between men and women in 1995. Now that we know there is a difference in salary between the two groups, let us examine further if sex bias exists. 

Linear Regression modeling salary as a function of sex, low percentage of variance explained by sex only, but statistically significant 
```{r}
lm_model <- lm(salary ~ sex, data = df_starting)
summary(lm_model)
```

Linear Regression modeling salary as a function of sex, accounting for  deg, field, rank, admin
```{r}
lm_model_2 <- lm(salary ~ sex + deg + field + rank + admin, data = df_starting)
summary(lm_model_2)
```

Linear Regression modeling salary as a function of sex, accounting for field, rank, admin, dist_from_degree
```{r}
lm_model_3 <- lm(salary ~ sex + field + rank + admin + dist_from_degree, data = df_starting)
summary(lm_model_3)
```

When you begin accounting for other variables, sex no longer is statistically significant to the model, interestingly. 

Check constant Variance Assumption part 1
```{r, echo=FALSE}
plot(lm_model_3$fitted.values, resid(lm_model_3),
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs. Fitted Values")
abline(h = 0, col = "red")
```
Check Constant Variance Assumption part 2
```{r, echo =FALSE}
qqnorm(resid(lm_model_3))
```

Check if interaction between sex and other significant variables produces any significance.
```{r}
lm_model_4 <- lm(salary ~ sex * field + sex * rank + admin + dist_from_degree, data = df_starting)
summary(lm_model_4)

```

Interaction terms when futher considering sex remain non-significant -- evidence that there is no discrimination taking place during start year. 

```{r}
ggplot(df_starting, aes(x = startyr, y = salary, color = sex)) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Starting Salary Trends Over Time by Sex", x = "Hiring Year", y = "Starting Salary")
```

Men's and Women's salaries, although different, do increase at a similar rate, further supporting that discrimination does not exist. 


```{r}
ggplot(df_starting, aes(x = rank, fill = sex)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Academic Rank by Sex", x = "Rank", y = "Count")
```

## Comparing Models 
```{r}
library(sjPlot)

tab_model(lm_model, lm_model_2, lm_model_3, lm_model_4,
          show.ci = FALSE,  # Hide confidence intervals
          show.std = TRUE,  # Show standardized coefficients
          show.se = TRUE,   # Show standard errors
          dv.labels = c("Starting Salary"))
```


### Disregard, Just Looking Into Something -- Not part of Project


Potentially a rank issue -- women are hired into associate or full positions far less frequently.
```{r}
df_starting_1 <- df_starting
df_starting_1$rank <- as.numeric(factor(df_starting$rank, levels = c("Assist", "Assoc", "Full"), ordered = TRUE))
```


```{r}
lm_rank <- lm(rank ~ startyr + dist_from_degree + sex * field, data = df_starting_1)
summary(lm_rank)
```



